---
title: "Class 11: AlphaFold"
author: "Madina Khorami (A18555185)"
format: pdf
toc: true
---

## The EBI Alphafold Database
T he EBI alphafold database contains alots of computed structure models. It is increasingly kely that the structure you are interested in is already in this database at https://alphafold.ebi.ac.uk
There are three major outputs from Alphafold
1. A model of structure in PDB.
2. A pIDDt score: that tells us how confident the model is for a given residue in your protein (high value is good, above 70)
3. A PAE score that tells us about protein packing quality


## Running AlphaFold
We will use Colabfold to run Alphafold on our sequence https://github.com/sokrypton/ColabFold?tab=readme-ov-file Figure from Alphafold here!



```{r}
library(bio3d)

pdb <- read.pdb("hivpr_dimer_23119/hivpr_dimer_23119_unrelaxed_rank_002_alphafold2_multimer_v3_model_1_seed_000.pdb")

pdb
```

Make a vector of input PDB file names that we can read into R.

```{r}
pdbfiles <- list.files("hivpr_dimer_23119/", pattern = ".pdb",
           full.names = T)
pdbfiles
```

```{r}
library (bio3dview)
pdbs <- pdbaln(pdbfiles, fit = TRUE, exefile = "msa")
```

```{r}
#| eval: !expr knitr::is_html_output()
view.pdbs(pdbs)
```


```{r}
# Quick view of model sequence
pdbs
```


```{r}
rd <- rmsd(pdbs, fit = TRUE)
```

```{r}
range(rd)
```

```{r}
dim(rd)
```


```{r}

library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
# Read a reference PDB Structure

pdb <- read.pdb("1HSG")
```
Obtain secondary strucutre from a call to `Stride()`, or `dssp()` on any of the model strucutres.

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

We can improve the superposition/fitting of our models by finding the most consistent **"rigid core"** common across all the models. For this we will use the core. `find()` function:

```{r}
core <- core.find(pdbs)
```

We can now use the identified core atom positions as a basis for a more suitable superposition and write out the fitted structures to a directory called corefit_structures:

```{r}
core.inds <- print(core, vol = 0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath = "corefit_structures")
```


Now we can examine the RMSF between positions of the structure. RMSF is an often used measure of conformational variance along the structure:

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```




## Predicted Alignment Error for Domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path="hivpr_dimer_23119/",
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```



```{r}
## For example purposes lets read the 1st and 5th files
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```




```{r}
# Per-residue pLDT scores

#same as B-factor of PDB..

head (pae1$plddt)
```


```{r}
pae1$max_pae
```


```{r}
pae5$max_pae
```


We can plot the N by N (where N is the number of residues) PAE scores with ggplot or with
functions from the Bio3D package:

```{r}
plot.dmat (pae1$pae,
xlab="Residue Position (i)",
ylab="Residue Position (j)")
```




```{r}
plot.dmat (pae5$pae,


xlab="Residue Position (i)",

ylab="Residue Position (j)",
grid.col = "black",

zlim=c(0,30))
```



```{r}
#We should really plot all of these using the same z range. Here is the model 1 plot again b
plot.dmat (pae1$pae,
xlab="Residue Position (i)", ylab="Residue Position (j)",
grid.col = "black",
zlim=c (0,30))
```



## Residue conservation from alignment file

```{r}
aln_file <- list.files(path="hivpr_dimer_23119/",
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the `conserv()` function.

```{r}
sim <- conserv(aln)
```


```{r}
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```
For a final visualization of these functionally important sites we can map this conservation score to the Occupancy column of a PDB file for viewing in molecular viewer programs such as Mol*, PyMol, VMD, chimera etc.

```{r}
m1.pdb <- read.pdb(pdbfiles[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```

