---
title: "Class 5: Data Viz with ggplot2"
author: "Madina(PID: 185)"
format: pdf
toc: true
---

## Background

There are many graphics system in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(cars)
```

Before I can use the **ggplot2** I need to install it on my computer.
To do this we can use the function `install.packages("ggplot2")`.

> **N.B.** We never run `install.packages()` in our quarto doc (we run it once only in R console) as it would re-install the package everytime we render the code.

Once installed we need to load up the package into our R brain
```{r}
library(ggplot2)
```


The main function in **ggplot2** package is called `ggplot()`
```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers

- The **data** (a data.frame of the stuff we wnat to plot)
- The **aes**thetics ( how the data maps to the plot)
- The **geom** layer(how you want the plot drawn, e.g points, line, etc.)

```{r}
ggplot(cars) +
  aes(x=speed ,y=dist ) + 
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between the speec and distance.

```{r}
ggplot(cars) +
  aes(x=speed ,y=dist ) + 
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed ,y=dist ) + 
  geom_point() +
  geom_smooth() +
  theme_bw() + 
  labs(title = "Stoping Distribution of Old Cars",
       x="Speed(MPH)",
       y="Distance(ft)")
```



> Q. Can you maek the `geom_smooth()` function produce alinea straight line fit to the data trend line and trun-off the grey error region.

```{r}
ggplot(cars) +
  aes(x=speed ,y=dist ) + 
  geom_point() +
  geom_smooth(method = lm, se=FALSE) +
  theme_bw() + 
  labs(title = "Stoping Distribution of Old Cars",
       x="Speed(MPH)",
       y="Distance(ft)")
```

---

## Gene Expression Data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


```{r}
up <- sum(genes$State == "up")
up
total <- nrow(genes)
total
ncol(genes)
fraction <- up/total
fraction
round(fraction, 2)
```
 A useful new function in this context is the `table()` function.
 
```{r}
table(genes$State)
```
 
 
My first plot attempt

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() +
  labs(title="Gene Expression Changes upon GLP-1 Inhibitor Drug Treatment", x="Control(no drug)", y="Drug Treatment") + 
  scale_color_manual(values = c("purple", "grey","orange"))
```


Here we read the famous gapminder

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)

```

> Q. How many entries (i.e. rows are in this dataset)

```{r}
nrow(gapminder)
```

> Q. How many "countries" are in this dataset?

```{r}
length(table(gapminder$country))
```

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset.

Plot of the "gdpPercap" vs "lifeExp" colored by "continent".

```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

```{r}
p
```


I can add more layers to "p"
```{r}
p + facet_wrap(~continent)
```

Make a plot for 1977 and 2007 not all year in the dataset.

First use the **dplyr** package and the `filter()` function from the package to extract the rows from the year 2007.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
gapminder_year_2007 <- filter(gapminder, year == 2007)
gapminder_year_1977 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year ==2007 | year == 1977)
```

```{r}
ggplot(gapminder_year_2007) + 
aes(gdpPercap, lifeExp, col=continent, size=pop) + 
  geom_point()
```

```{r}
ggplot(g) + 
  aes(gdpPercap, lifeExp, col=continent, size = pop) + 
  geom_point() + 
  facet_wrap(~year)

```

> Q. Make a histogram of lifeExp colored by continent.(use `fill=continent` or `col=continent`)


```{r}
ggplot(gapminder) + 
  aes(lifeExp, fill = continent) + 
  geom_histogram()
```

> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) + 
  aes(lifeExp, fill = continent) + 
  geom_histogram() + 
  facet_wrap(~continent)
```
