---
title: "Class 7: Machine Learning 1"
author: "Madina Khorami (A18555185)"
format: pdf
toc: TRUE
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include, **clustering** and **dimensionallity reductiontio**.

## K-means Clustering

The main function in "base" R for k-mean clustering is called `kmeans()`.

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To help us do this we can use the `rnowm()` function:


```{r}
# The default mean is at 0 but if we give a value for mean it would center the data around that value

hist( rnorm(10000, mean = 3) )
```



```{r}
x <- c(rnorm(30, mean = -3), rnorm(30, mean = 3))

```

```{r}
# Reverse the order of the value defined for x, rev(x)
# cbind() are columnize the vector and rbind() in rows
z <- cbind(x = x, y = rev(x))
plot(z)
```



Now we can run `kmeans()` on this input `z` to see what the results look like.

```{r}

## centers tells how many cluster do we want and the cluster vector in the result says what cluseter each value depends on so it would be 1 or 2.

km <- kmeans(z, centers = 2)
km
```



```{r}
## attributes store information of kmeans that are not primary data more like extra info.
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```


> Q. What "component of your result object details cluster assignment/membership?

```{r}
km$cluster
```



> Q. What "component of your result object details cluster center"?

```{r}
km$centers
```

> Q. Plot`z` colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
## color will be assigned to like one red 2 is blue 3 is red ....

plot(z, col = c("red","blue"))

```


```{r}
## As each number represents a color we can just type in the number and the color will show up
## Also we can assign the cluster of k mean as color and it will perfectly devide them based on their cluster.
plot(z, col = km$cluster)
points(km$centers, col = "blue", pch = 15)
```

> Q. Run a kmeans clustering and add plot the results asking for 4 clusters (K=4)?


```{r}
km4 <- kmeans(z, centers = 4)
plot(z, col=km4$cluster)
points(km4$centers, col = km4$cluster, pch = 15)

```



> **N.B.** You need to tell K-means the number of clusters (i.e set `centers=2`)!!


One approach is to try different values for `centers` and then pick the best ...

```{r}
## tot.withinss give u the center value 
ans <- NULL
for (i in 1:10) {
  km <- kmeans(z, centers = i)
ans <- c(ans, km$tot.withinss)

}
plot(ans,  typ ="o",
     xlab=" Number of Clusters", ylab= "Total Number of Within SS")


```


## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`.

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`.

```{r}
# first build the distance matrix where it print out the result symmetrical.
d <- dist(z)

## now create the hierachical clustering of the matrix
hc <- hclust(d)
hc
```


There is a bespoke `plot()` method for `hclust()` result object.

```{r}
plot(hc)
abline(h=8, col = "red")
```

Once we have our `hclust()` object (Our "tree of cluster dendogram") we can *"cut"* the tree to reval the clustering pattern.

```{r}
cutree(hc, h=8)
```


> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)


```{r}
grps <- cutree(hc, k=2)
plot(z, col = grps)
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction method that is popular for revealing patterns in complex datasets.

## Analysis of UK Food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that hace some regions being distinct from others.

## Data Import

The data is made available in csv format so we can use the `read.csv()` function to read the data.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
nrow(x)
ncol(x)
```

> Preview the first 6 rows

```{r}
head(x, n=6)
```


> Note how the minus indexing works

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```
 
 
 > To check the Dimension
 
```{r}
dim(x)
```
  
> Another way to change the row name of the dataset

```{r}
x <- read.csv(url, row.names=1)
head(x)
```
  
  > Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
  
  I prefer the second approach `read.csv(url, rownames...)` because it is easy to use and error-prone, as it introduce the row name at the time we import the file so no need to take another step for introducing a new object. This way we can also avoid forgetting to remove the row name later on since it is done at the beginning. whereas manually setting `rownames(x) <- x[,1]` works but is easier to mess up, especially with messy or inconsistent data.
  
  
```{r}
# Using base R
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
  
  > Q3: Changing what optional argument in the above barplot() function results in the following plot? 
  
  By changeing the beside to False we tell the barplot to create stacks the rows on top of each other into a singular verticlebar and this is the default as well.
  
```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
  

## Tidy data
fix anything that went wrong with data import

```{r}
# Currently we have wide format
dim(x)
```
```{r}
head(x)

```

To convert this to “long” format we want one row per measurement - maximizes rows**&(17x4=68)**, minimizes columns (with a singe Consumption measurement value per Country). We will do tidying with the `pivot_longer()` function from the tidyr package:

```{r}

library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```

```{r}
head(x_long)

```


```{r}

library(ggplot2)
```

```{r}
# Create grouped bar plot

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

> Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

To make the plot we change the position in`geom_col()` to default which is stacked bars of rows. so remove the position from dodge to default.

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col() +
  theme_bw()
```


## Exploratory Analysis
Make some plots to help make sense of obvious trends.

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?


```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

The `pairs()` code creates a matrix of scatterplots showing all pairwise comparisons between the variables in x, with each point representing a country and colored uniquely.

If a point lies on the diagonal in any of the off-diagonal plots, it means that for that country the two variables being compared have equal or very similar values (so x = y for that pair).


## Heatmap: Another Data Visualization

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

**Key-point** Even relatively small datasets can prove challenging to interpret

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on both the pairs plot and the heatmap, England, Wales, and Scotland cluster closely together, while Northern Ireland appears more separate, suggesting that England, Wales, and Scotland have more similar overall food consumption patterns, whereas Northern Ireland differs somewhat. The heatmap hints that Northern Ireland tends to have different levels in certain categories (like cereals, potatoes, or fats) compared to the others.


## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function expects to be rows and the "variables" to be columns.

So here we need to take the traspose of our `x` input object.

```{r}
## t will transpose the data, this will tell how well the capture of the component is by using the prcomp
pca <- prcomp( t(x) )
summary(pca)
```

The returned `pca` object has components that wr can use to make out main result figure.
PC1 is more important because it has more varience than all other PCs.

```{r}
attributes(pca)
```


The main result figure from this analysis is called a "PC score plot" or " Ordenation plot", "PC plot" or PC1 vs PC2.


```{r}
pca$x

```


> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.


```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```


> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
plot_df <- data.frame(pca$x)
plot_df$Country <- rownames(plot_df)

ggplot(plot_df, aes(x = PC1, y = PC2)) +
   geom_point(color = "grey60",size = 2) +    
  geom_text(aes(label = Country, color = Country), size = 3, vjust = -1) +
  scale_color_manual(values = c(
    "England" = "orange",
    "Wales" = "red",
    "Scotland" = "blue",
    "N.Ireland" = "darkgreen"
  )) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw() +
  theme(legend.position = "none")
```


> Below we can use the square of pca$sdev , which stands for “standard deviation”, to calculate how much variation in the original data each PC accounts for.

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
## or the second row here...
z <- summary(pca)
z$importance
```

```{r}
# Create scree plot with ggplot
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```


```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```
Here we see observations (foods) with the largest positive loading scores that effectively “push” N. Ireland to right positive side of the plot `(including Fresh_potatoes and Soft_drinks)`.

We can also see the observations/foods with high negative scores that push the other countries to the left side of the plot `(including Fresh_fruit and Alcoholic_drinks)`.



> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

```


> The two food groups that usually stand out most on *PC2* are **Fresh_fruit** and **Fresh_potatoes**. *PC2* mainly tells us about a second pattern of variation that separates countries based on differences in specific food preferences, especially fresh produce consumption, rather than the overall general consumption level that *PC1* captures.


**Key-Point**: PCA basically helps simplify the data by reducing it from 17 variables down to just 2 main components, while still keeping most of the important information. From this, we can see that England, Wales, and Scotland behave pretty similarly, while Northern Ireland stands out as being different. This makes it easier to visualize patterns and compare countries without getting overwhelmed by too many variables. It also shows that most of the variation in the data can be explained by just a couple of key trends.
















